---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd instead. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# survAHadjust

DOI: [10.5281/zenodo.20277580](https://doi.org/10.5281/zenodo.20277580)

`survAHadjust` calculates average hazard-based summaries from adjusted survival curves.

The main workflow is:

```text
adjusted survival curves
  -> adjusted_ahsw()
  -> ahsw_table()
```

For most adjusted survival curve methods, users can first use `adjustedCurves::adjustedsurv()`.

For AIPTW, `survAHadjust` provides `aiptw_adjustedsurv()`.

Both workflows produce an adjusted survival curve object that can be passed to `adjusted_ahsw()`.

## Installation

Install the GitHub version with:

```r
install.packages("pak")
pak::pak("kkevin821/survAHadjust")
```

For local development, after cloning the repository and opening the package folder in RStudio, install the package locally with:

```r
devtools::install()
```

## Basic example

`adjusted_ahsw()` requires an adjusted survival curve object with:

```text
adjsurv$adj$time
adjsurv$adj$surv
adjsurv$adj$group
```

```{r basic-example}
library(survAHadjust)

adjsurv <- list(
  adj = data.frame(
    time = c(0, 0.5, 1.0, 0, 0.5, 1.0),
    group = factor(c(0, 0, 0, 1, 1, 1)),
    surv = c(1.0, 0.80, 0.70, 1.0, 0.90, 0.85)
  )
)

res <- adjusted_ahsw(adjsurv, to = 1, conf_int = FALSE)

res$est
```

The result can be formatted as a one-row table:

```{r table-example}
tab_example <- ahsw_table(res, method = "Example")

knitr::kable(
  tab_example,
  align = c("l", "c", "c", "c", "c")
)
```

For a cleaner display in R Markdown, Quarto, or pkgdown pages, users can pass the output from `ahsw_table()` to `knitr::kable()`.

## Standard adjusted survival curve methods

For standard adjusted survival curve methods, use `adjustedCurves::adjustedsurv()` first.

For example:

```r
outcome_model <- survival::coxph(
  survival::Surv(time, event) ~ group + x1 + x2 + x3,
  data = dat,
  x = TRUE
)

adjsurv <- adjustedCurves::adjustedsurv(
  data = dat,
  variable = "group",
  ev_time = "time",
  event = "event",
  method = "direct",
  outcome_model = outcome_model,
  bootstrap = TRUE,
  n_boot = 300
)

res <- adjusted_ahsw(
  adjsurv = adjsurv,
  to = 0.7,
  conf_int = TRUE
)

ahsw_table(res, method = "Direct standardization")
```

Other methods from `adjustedCurves::adjustedsurv()` can be used in the same way.

## AIPTW

AIPTW is a standard doubly robust approach for estimating adjusted survival
curves. In `survAHadjust`, the function `aiptw_adjustedsurv()` implements the
AIPTW estimator used in our simulation study and application. The implementation
follows the AIPTW/IPCW setup described by Ozenne et al. (2020). Under
independent censoring, the censoring distribution is estimated using a pooled
Kaplan-Meier estimator; when `covars_cens` is supplied, a Cox model is used for
the censoring distribution.

For non-AIPTW adjusted survival curve methods, users can use
`adjustedCurves::adjustedsurv()` and then pass the resulting object to
`adjusted_ahsw()`.

If users prefer another implementation of AIPTW, they can also use that output
with `adjusted_ahsw()` as long as it is formatted as an adjusted survival curve
object with an `adj` element containing columns `time`, `surv`, and `group`.

For bootstrap-based confidence intervals, the object should also contain
`boot_data` with columns `time`, `surv`, `group`, and `boot`.

```r
adjsurv_aiptw <- aiptw_adjustedsurv(
  data = dat,
  covars_treat = c("x2", "x3", "x5", "x6"),
  covars_outcome = c("x1", "x2", "x4", "x5_squared"),
  covars_cens = NULL,
  y_col = "time",
  delta_col = "event",
  z_col = "group",
  bootstrap = TRUE,
  n_boot = 300
)

res_aiptw <- adjusted_ahsw(
  adjsurv = adjsurv_aiptw,
  to = 0.7,
  conf_int = TRUE
)

ahsw_table(res_aiptw, method = "AIPTW")
```

If censoring depends on covariates, specify `covars_cens`:

```r
adjsurv_aiptw <- aiptw_adjustedsurv(
  data = dat,
  covars_treat = c("x2", "x3", "x5", "x6"),
  covars_outcome = c("x1", "x2", "x4", "x5_squared"),
  covars_cens = c("x2"),
  y_col = "time",
  delta_col = "event",
  z_col = "group",
  bootstrap = TRUE,
  n_boot = 300
)
```

## Output

`adjusted_ahsw()` always returns point estimates in:

```r
res$est
```

The object `res$est` contains RMST, event rate, and AHSW:

```text
              group0  group1  difference  log_dif
rmst
event_rate
ahsw
```

Here, `difference` is:

```text
group1 - group0
```

and `log_dif` is:

```text
log(group1) - log(group0)
```

For the AHSW row, `exp(log_dif)` gives the ratio of average hazards.

If `conf_int = TRUE` and bootstrap data are available, `adjusted_ahsw()` also returns bootstrap-based inference tables:

```r
res$rmst
res$evrt
res$ahsw
```

These tables give inference results for RMST, event rate, and AHSW, respectively. Each table has rows:

```text
group0
group1
difference
log_dif
```

and columns:

```text
est       point estimate
se        bootstrap standard error
low_XX    lower confidence limit
high_XX   upper confidence limit
p_value   two-sided p-value
```

Here, `XX` is determined by the user-specified `conf_level`. For example, when `conf_level = 0.95`, the confidence interval columns are:

```text
low_95
high_95
```

When `conf_level = 0.90`, the confidence interval columns are:

```text
low_90
high_90
```

For example, with `conf_level = 0.90`, the AHSW inference table has the following structure:

```text
                   est       se     low_90   high_90   p_value
group0
group1
difference
log_dif
```

The helper function `ahsw_table()` uses the confidence interval level in the column names. For example, with `conf_level = 0.90`, the formatted table uses column names such as:

```text
Group 0 AH (90% CI)
Group 1 AH (90% CI)
DAH (90% CI)
RAH (90% CI)
```

The `p_value` is reported for `difference` and `log_dif`. The p-values for `group0` and `group1` are set to `NA`, because the main comparisons are the difference and ratio between groups.

The `log_dif` row is the log ratio of average hazards. When `conf_level = 0.95`, the ratio and its confidence interval can be obtained by exponentiating the log-scale values:

```r
exp(res$ahsw["log_dif", c("est", "low_95", "high_95")])
```

If a different confidence level is used, replace `low_95` and `high_95` with the corresponding column names.

A worked example with bootstrap-based confidence intervals, including `low_90`, `high_90`, and table columns such as `Group 0 AH (90% CI)`, is provided in the vignette.

## Documentation

Package documentation, function references, and the vignette are available at:

https://kkevin821.github.io/survAHadjust/

The vignette can also be opened directly at:

https://kkevin821.github.io/survAHadjust/articles/survAHadjust.html

The vignette source file is available in:

```text
vignettes/survAHadjust.Rmd
```

## Citation

If you use `survAHadjust`, please cite the software DOI from Zenodo:

```text
Xiong H, Uno H. survAHadjust: Adjusted Average Hazard with Survival Weight. Zenodo. https://doi.org/10.5281/zenodo.20277580
```
