---
title: "Using survAHadjust"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using survAHadjust}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

The `survAHadjust` package calculates average hazard-based summaries from adjusted survival curves.

The main workflow is:

```text
adjusted survival curves
  -> adjusted_ahsw()
  -> ahsw_table()
```

For most adjusted survival curve methods, users can use `adjustedCurves::adjustedsurv()`.

For AIPTW, this package provides `aiptw_adjustedsurv()`.

# Load packages

```{r setup}
library(survAHadjust)
library(survival)
```

# Example data

We use the `pbc` data from the `survival` package only as a small example dataset.

```{r data}
D <- survival::pbc[!is.na(survival::pbc$trt), ]

D$Y <- D$time / 365.25
D$Delta <- as.numeric(D$status == 2)
D$Z <- as.numeric(D$trt == 2)

D <- D[complete.cases(D[, c("Y", "Delta", "Z", "age", "sex", "bili", "albumin")]), ]

head(D[, c("Y", "Delta", "Z", "age", "sex", "bili", "albumin")])
```

# Workflow 1: Analyze an existing adjusted survival curve object

The function `adjusted_ahsw()` accepts an object that contains adjusted survival curves in `adjsurv$adj`.

The required columns are:

```text
time
surv
group
```

```{r fake-adjsurv}
adjsurv_example <- list(
  adj = data.frame(
    time = c(0, 0.5, 1.0, 0, 0.5, 1.0),
    group = factor(c(0, 0, 0, 1, 1, 1)),
    surv = c(1.0, 0.80, 0.70, 1.0, 0.90, 0.85)
  )
)

res_example <- adjusted_ahsw(adjsurv_example, to = 1, conf_int = FALSE)

res_example$est
```

The result can be formatted using `ahsw_table()`.

```{r fake-table}
tab_example <- ahsw_table(res_example, method = "Example")

knitr::kable(
  tab_example,
  align = c("l", "c", "c", "c", "c")
)
```

# Workflow 2: Standard methods from adjustedCurves

For standard adjusted survival curve methods, users can first call `adjustedCurves::adjustedsurv()` and then pass the result to `adjusted_ahsw()`.

The following code shows the workflow. It is not evaluated in this vignette because it depends on `adjustedCurves` and its additional dependencies.

```{r adjustedcurves-direct, eval = FALSE}
D_ac <- D
D_ac$Z <- factor(D_ac$Z)

outcome_model <- survival::coxph(
  survival::Surv(Y, Delta) ~ Z + age + sex + bili + albumin,
  data = D_ac,
  x = TRUE
)

adjsurv_direct <- adjustedCurves::adjustedsurv(
  data = D_ac,
  variable = "Z",
  ev_time = "Y",
  event = "Delta",
  method = "direct",
  outcome_model = outcome_model,
  conf_int = FALSE,
  bootstrap = FALSE
)

res_direct <- adjusted_ahsw(
  adjsurv = adjsurv_direct,
  to = 5,
  conf_int = FALSE
)

tab_direct <- ahsw_table(res_direct, method = "Direct standardization")

knitr::kable(
  tab_direct,
  align = c("l", "c", "c", "c", "c")
)
```

Other methods from `adjustedCurves::adjustedsurv()` can be used in the same way, such as IPTW, matching, empirical likelihood, and Kaplan-Meier methods.

# Workflow 3: AIPTW

AIPTW is a standard doubly robust approach for estimating adjusted survival
curves. In `survAHadjust`, the function `aiptw_adjustedsurv()` implements the
AIPTW estimator used in our simulation study and application. The implementation
follows the AIPTW/IPCW setup described by Ozenne et al. (2020). Under
independent censoring, the censoring distribution is estimated using a pooled
Kaplan-Meier estimator; when `covars_cens` is supplied, a Cox model is used for
the censoring distribution.

For non-AIPTW adjusted survival curve methods, users can use
`adjustedCurves::adjustedsurv()` and then pass the resulting object to
`adjusted_ahsw()`.

If users prefer another implementation of AIPTW that returns an adjusted
survival curve object with an `adj` element containing columns `time`, `surv`,
and `group`, that object can also be passed to `adjusted_ahsw()`.

```{r aiptw}
adjsurv_aiptw <- aiptw_adjustedsurv(
  data = D,
  covars_treat = c("age", "sex"),
  covars_outcome = c("bili", "age", "albumin"),
  y_col = "Y",
  delta_col = "Delta",
  z_col = "Z",
  bootstrap = FALSE
)

head(adjsurv_aiptw$adj)
```

Then pass the result to `adjusted_ahsw()`.

```{r aiptw-ahsw}
res_aiptw <- adjusted_ahsw(
  adjsurv = adjsurv_aiptw,
  to = 5,
  conf_int = FALSE
)

res_aiptw$est
```

Format the result:

```{r aiptw-table}
tab_aiptw <- ahsw_table(res_aiptw, method = "AIPTW")

knitr::kable(
  tab_aiptw,
  align = c("l", "c", "c", "c", "c")
)
```

# Bootstrap inference

Bootstrap inference can be requested by setting `bootstrap = TRUE`.

For real analyses, use a larger number of bootstrap samples, such as `n_boot = 300`. The following code is not evaluated in this vignette because it can take longer to run.

```{r bootstrap-example, eval = FALSE}
adjsurv_aiptw_boot <- aiptw_adjustedsurv(
  data = D,
  covars_treat = c("age", "sex"),
  covars_outcome = c("bili", "age", "albumin"),
  y_col = "Y",
  delta_col = "Delta",
  z_col = "Z",
  bootstrap = TRUE,
  n_boot = 300
)

res_aiptw_boot <- adjusted_ahsw(
  adjsurv = adjsurv_aiptw_boot,
  to = 5,
  conf_int = TRUE
)

tab_aiptw_boot <- ahsw_table(res_aiptw_boot, method = "AIPTW")

knitr::kable(
  tab_aiptw_boot,
  align = c("l", "c", "c", "c", "c")
)
```

The code above shows the workflow for real analyses. The small example below is evaluated to show the structure of the bootstrap-based output and how the confidence interval level appears in the output column names.

```{r ci-level-example}
adjsurv_boot_example <- list(
  adj = data.frame(
    time = c(0, 0.5, 1.0, 0, 0.5, 1.0),
    group = factor(c(0, 0, 0, 1, 1, 1)),
    surv = c(1.0, 0.80, 0.70, 1.0, 0.90, 0.85)
  ),
  boot_data = rbind(
    data.frame(
      time = c(0, 0.5, 1.0, 0, 0.5, 1.0),
      group = factor(c(0, 0, 0, 1, 1, 1)),
      surv = c(1.0, 0.82, 0.72, 1.0, 0.91, 0.86),
      boot = 1
    ),
    data.frame(
      time = c(0, 0.5, 1.0, 0, 0.5, 1.0),
      group = factor(c(0, 0, 0, 1, 1, 1)),
      surv = c(1.0, 0.78, 0.68, 1.0, 0.89, 0.84),
      boot = 2
    )
  )
)

res_boot_90 <- adjusted_ahsw(
  adjsurv = adjsurv_boot_example,
  to = 1,
  conf_int = TRUE,
  conf_level = 0.90
)

res_boot_90$ahsw
```

When `conf_level = 0.90`, the confidence interval columns are named `low_90` and `high_90`. Similarly, when `conf_level = 0.95`, the columns are named `low_95` and `high_95`.

The helper function `ahsw_table()` uses these confidence interval column names to label the summary table. For example, because the object above was created with `conf_level = 0.90`, the table columns include `Group 0 AH (90% CI)`, `Group 1 AH (90% CI)`, `DAH (90% CI)`, and `RAH (90% CI)`.

The helper function `ahsw_table()` returns a regular data frame. In rendered documents, `knitr::kable()` can be used to display it as a cleaner table. The `<br>` tags below control line breaks in the column names.

```{r ci-level-table-example}
tab_boot_90 <- ahsw_table(res_boot_90, method = "Example")

knitr::kable(
  tab_boot_90,
  align = c("l", "c", "c", "c", "c"),
  col.names = c(
    "Method",
    "Group 0 AH<br>(90% CI)",
    "Group 1 AH<br>(90% CI)",
    "DAH<br>(90% CI)",
    "RAH<br>(90% CI)"
  ),
  escape = FALSE
)
```

# Combining multiple methods

Results from multiple methods can be combined with `rbind()`.

The following code shows how to combine results from standard methods and AIPTW. It is not evaluated in this vignette because `res_direct` is created in a non-evaluated example above.

```{r combine-multiple-methods, eval = FALSE}
tab_all <- rbind(
  ahsw_table(res_direct, method = "Direct standardization"),
  ahsw_table(res_aiptw, method = "AIPTW")
)

knitr::kable(
  tab_all,
  align = c("l", "c", "c", "c", "c")
)
```

For the evaluated example in this vignette, we only show the AIPTW row.

```{r combine-aiptw-only}
tab_aiptw_only <- rbind(
  ahsw_table(res_aiptw, method = "AIPTW")
)

knitr::kable(
  tab_aiptw_only,
  align = c("l", "c", "c", "c", "c")
)
```
