---
title: "Using survAHadjust"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using survAHadjust}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

The `survAHadjust` package calculates average hazard-based summaries from adjusted survival curves.

The main workflow is:

```text
adjusted survival curves
  -> adjusted_ahsw()
  -> ahsw_table()
```

For most adjusted survival curve methods, users can use `adjustedCurves::adjustedsurv()`.

For AIPTW, this package provides `aiptw_adjustedsurv()`.

# Load packages

```{r setup}
library(survAHadjust)
library(survival)
```

# Example data

We use the `pbc` data from the `survival` package only as a small example dataset.

```{r data}
D <- survival::pbc[!is.na(survival::pbc$trt), ]

D$Y <- D$time / 365.25
D$Delta <- as.numeric(D$status == 2)
D$Z <- as.numeric(D$trt == 2)

D <- D[complete.cases(D[, c("Y", "Delta", "Z", "age", "sex", "bili", "albumin")]), ]

head(D[, c("Y", "Delta", "Z", "age", "sex", "bili", "albumin")])
```

# Workflow 1: Analyze an existing adjusted survival curve object

The function `adjusted_ahsw()` accepts an object that contains adjusted survival curves in `adjsurv$adj`.

The required columns are:

```text
time
surv
group
```

```{r fake-adjsurv}
adjsurv_example <- list(
  adj = data.frame(
    time = c(0, 0.5, 1.0, 0, 0.5, 1.0),
    group = factor(c(0, 0, 0, 1, 1, 1)),
    surv = c(1.0, 0.80, 0.70, 1.0, 0.90, 0.85)
  )
)

res_example <- adjusted_ahsw(adjsurv_example, to = 1, conf_int = FALSE)

res_example$est
```

The result can be formatted using `ahsw_table()`.

```{r fake-table}
ahsw_table(res_example, method = "Example")
```

# Workflow 2: Standard methods from adjustedCurves

For standard adjusted survival curve methods, users can first call `adjustedCurves::adjustedsurv()` and then pass the result to `adjusted_ahsw()`.

The following code shows the workflow. It is not evaluated in this vignette because it depends on `adjustedCurves` and its additional dependencies.

```{r adjustedcurves-direct, eval = FALSE}
D_ac <- D
D_ac$Z <- factor(D_ac$Z)

outcome_model <- survival::coxph(
  survival::Surv(Y, Delta) ~ Z + age + sex + bili + albumin,
  data = D_ac,
  x = TRUE
)

adjsurv_direct <- adjustedCurves::adjustedsurv(
  data = D_ac,
  variable = "Z",
  ev_time = "Y",
  event = "Delta",
  method = "direct",
  outcome_model = outcome_model,
  conf_int = FALSE,
  bootstrap = FALSE
)

res_direct <- adjusted_ahsw(
  adjsurv = adjsurv_direct,
  to = 5,
  conf_int = FALSE
)

ahsw_table(res_direct, method = "Direct standardization")
```

Other methods from `adjustedCurves::adjustedsurv()` can be used in the same way, such as IPTW, matching, empirical likelihood, and Kaplan-Meier methods.

# Workflow 3: AIPTW

AIPTW is a standard doubly robust approach for estimating adjusted survival
curves. In `survAHadjust`, the function `aiptw_adjustedsurv()` implements the
AIPTW estimator used in our simulation study and application. The implementation
follows the AIPTW/IPCW setup described by Ozenne et al. (2020). Under
independent censoring, the censoring distribution is estimated using a pooled
Kaplan-Meier estimator; when `covars_cens` is supplied, a Cox model is used for
the censoring distribution.

For non-AIPTW adjusted survival curve methods, users can use
`adjustedCurves::adjustedsurv()` and then pass the resulting object to
`adjusted_ahsw()`.

If users prefer another implementation of AIPTW that returns an adjusted
survival curve object with `adj`, `time`, `surv`, and `group`, that object can
also be passed to `adjusted_ahsw()`.

```{r aiptw}
adjsurv_aiptw <- aiptw_adjustedsurv(
  data = D,
  covars_treat = c("age", "sex"),
  covars_outcome = c("bili", "age", "albumin"),
  y_col = "Y",
  delta_col = "Delta",
  z_col = "Z",
  bootstrap = FALSE
)

head(adjsurv_aiptw$adj)
```

Then pass the result to `adjusted_ahsw()`.

```{r aiptw-ahsw}
res_aiptw <- adjusted_ahsw(
  adjsurv = adjsurv_aiptw,
  to = 5,
  conf_int = FALSE
)

res_aiptw$est
```

Format the result:

```{r aiptw-table}
ahsw_table(res_aiptw, method = "AIPTW")
```

# Bootstrap inference

Bootstrap inference can be requested by setting `bootstrap = TRUE`.

For real analyses, use a larger number of bootstrap samples, such as `n_boot = 300`. The following code is not evaluated in this vignette because it can take longer to run.

```{r bootstrap-example, eval = FALSE}
adjsurv_aiptw_boot <- aiptw_adjustedsurv(
  data = D,
  covars_treat = c("age", "sex"),
  covars_outcome = c("bili", "age", "albumin"),
  y_col = "Y",
  delta_col = "Delta",
  z_col = "Z",
  bootstrap = TRUE,
  n_boot = 300
)

res_aiptw_boot <- adjusted_ahsw(
  adjsurv = adjsurv_aiptw_boot,
  to = 5,
  conf_int = TRUE
)

ahsw_table(res_aiptw_boot, method = "AIPTW")
```

The example above shows the workflow for real analyses. The small example below is evaluated to show the structure of the bootstrap output and how the confidence interval level appears in the output column names.

```{r ci-level-example}
adjsurv_boot_example <- list(
  adj = data.frame(
    time = c(0, 0.5, 1.0, 0, 0.5, 1.0),
    group = factor(c(0, 0, 0, 1, 1, 1)),
    surv = c(1.0, 0.80, 0.70, 1.0, 0.90, 0.85)
  ),
  boot_data = rbind(
    data.frame(
      time = c(0, 0.5, 1.0, 0, 0.5, 1.0),
      group = factor(c(0, 0, 0, 1, 1, 1)),
      surv = c(1.0, 0.82, 0.72, 1.0, 0.91, 0.86),
      boot = 1
    ),
    data.frame(
      time = c(0, 0.5, 1.0, 0, 0.5, 1.0),
      group = factor(c(0, 0, 0, 1, 1, 1)),
      surv = c(1.0, 0.78, 0.68, 1.0, 0.89, 0.84),
      boot = 2
    )
  )
)

res_boot_90 <- adjusted_ahsw(
  adjsurv = adjsurv_boot_example,
  to = 1,
  conf_int = TRUE,
  conf_level = 0.90
)

res_boot_90$ahsw
```

When `conf_level = 0.90`, the confidence interval columns are named `low_90` and `high_90`. Similarly, when `conf_level = 0.95`, the columns are named `low_95` and `high_95`.

The helper function `ahsw_table()` uses these confidence interval column names to label the summary table.

```{r ci-level-table-example}
ahsw_table(res_boot_90, method = "Example")
```

# Combining multiple methods

Results from multiple methods can be combined with `rbind()`.

The following code shows how to combine results from standard methods and AIPTW. It is not evaluated in this vignette because `res_direct` is created in a non-evaluated example above.

```{r combine-multiple-methods, eval = FALSE}
tab_all <- rbind(
  ahsw_table(res_direct, method = "Direct standardization"),
  ahsw_table(res_aiptw, method = "AIPTW")
)

tab_all
```

For the evaluated example in this vignette, we only show the AIPTW row.

```{r combine-aiptw-only}
tab_aiptw <- rbind(
  ahsw_table(res_aiptw, method = "AIPTW")
)

tab_aiptw
```
